Does stomatal patterning in amphistomatous leaves minimize the CO\(_2\) diffusion path length within leaves?

### 2024-01-21 ms/response-to-reviewers.qmd

**Response:** \textcolor{blue}{We'’'ve added the relevant text to the Introduction on lines 54-73 as described in a previous response to another comment. We also added to the Discussion on lines 215-218:}

> "However, reducing this distance to its absolute minimum may be constrained by developmental processes or the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e. the selection coefficient is less than drift barrier *sensu* @sung_drift-barrier_2012). We also consider that our definition of optimal may be incorrect because it is based on overly simplistic assumptions about leaf mesophyll structure."

L252-253. Have alternative ideal distributions been tested, considering leaf structures like vein density, trichomes and substomatal cavity volume? Excluding these factors in estimations seems to be a significant omission. It seems problematic that other important leaf features, such as tortuosity, are not considered in the estimations but are discussed in the text. Again, I don't think the authors need to account for them, but please clarify the limitations of not doing so.

**Response:** \textcolor{blue}{To the best of our knowledge, no one has quantitatively investigated the optimal pattern with more realistic mesophyll structures. Most of the literature we are aware of predates the deeper understanding of heterogenous mesophyll structure that we have today. Our study highlights the inaccuracy of some predictions based on these simplisitic assumptions and will ophefully motivate revisiting these predictions in light of better empirical information. We added a sentence in the penultimate paragraph of the manuscript on lines 289-291 that addresses this point:}

> "To understand why stomata are not ideally dispersed, more modeling (with more realistic assumptions including vein density and IAS structure) should be done to estimate the photosynthetic properties of varying stomatal patterning."

## References
